Investigating Neolithic pottery use from Guijiabao in the Yanyuan Basin in southwestern China using organic residue analysis

### 2023-04-16 analysis/paper/paper.Rmd
edit the abstract and keywords

---
title: "Investigating Neolithic pottery use from Guijiabao in the Yanyuan Basin in southwest China using organic residue analysis"
author:
  - Li-Ying Wang:
      email: liyingw@asihp.net
      institute: [AcS]
      correspondence: true
  - Kuei-chen Lin:
      email: kueichenlin@asihp.net
      institute: [AcS]
      correspondence: false
  - Zhiqing Zhou:
      email: zhouzhiQ2000@163.com
      institute: [InsC]
      correspondence: false
institute:
  - AcS: Institute of History and Philology, Academia Sinica 
  - InsC: Chengdu Institute of Cultural Relics and Archaeology
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
    The spread of new crops or terrestial animals across areas in southwest China during the Neolithic period is one of the important archaeological issues to understand the process of adoption of new culinary practices in this region. It is widely accepted that rice was introduced into southwest China in 4700 BP at the latest and milletes came even earlier. However, to what extent new crops were used by local peoples at that time remains unclear. The excavation of Guijiabao in the Yanyuan basin in southwest China uncovered abundant materials with well-preserved stratigraphic records and has provided a unique opportunity to understand the local adoption of new crops. To further understand pottery use and food practices at Guijiabao, we investigated organic residues preserved in potsherds using lipid biomarker identification and compounds-specific stable isotope analysis. The results show signals of terrestrial animals in most potsherds and a C3 plant source in one potshard, during the Neolithic period. Our case study indicates that hunting might have been the primary means for obtaining subsistence, supplemented by rice/millet farming. This study also demonstrates an approach using organic residue analysis to explore pottery use and culinary practices during the Neolithic period in southwest China. A more systematic sampling of different types of potsherds and temporal phases with larger samples is needed to provide a clearer picture of ancient diets and subsistence.

keywords: |
  Lipid residue analysis; Ceramics; Southwest China; Stable isotope analysis; Food practices
highlights: |
  These are the highlights. 
---

@@

# Introduction
<!-- intro to the topic and research aim-->
Movements of people between geographically distinct areas are usually accompanied by the exchange of goods, technology, or ideas that could have cultural or economic impacts on local people groups involved in the interactions [@Burmeister2000]. The Hengduan Mountain ranges and valleys in southwestern China are geographical areas that connect the Qinghai-Tibetan Plateau and the Yunnan–Guizhou Plateau and by which movement to these plateaus must occur. They play an important role in the cultural exchanges and spread of crops into southwestern China and even further to mainland Southeast Asia [@Ma2021; @Lin2022; @Hao2022; @Tong1986]. Despite a long-standing debate on the routes of rice/millet dispersal from northwest China and the middle Yangtze valley to southwestern China [@Guedes2013; @DAlpoim2018; @Liu2022], it is widely accepted that mixed farming was practiced in southwest Sichuan around 4700 BP and in western Yunnan before 4600 BP [@Jiang2016; @YunnanMuseum1981; @Hao2022; @Zhang2010; @Dal2022; @Chen2015]. The Hengduan Mountain regions in the Neolithic period are therefore essential for understanding the processes of agricultural expansion. Recent archaeological evidence from the Yanyuan Basin in the middle Hengduan Mountains further reveals that millet and rice were present at roughly the same period as the grains found in Sichuan [@Huan2022; @Hao2022]. However, to what extent the two crops were incorporated into the local diet is still not fully understood.

@@

```{r SW-china-map, fig.cap = "The archaeological sites Guijiabao and Gaoshan and other locations in southwest China that are mentioned in this study. Map data is from naturalearthdata.com"}
library(here)
knitr::include_graphics(here("analysis", "figures", "Sichuan-sites-map.png"))
```

@@

Although the recent findings at the Guijiabao and Gaoshan sites support mixed agriculture as the primary subsistence strategy, it is still unclear how much millet or rice was included in the local diets during the Neolithic period. The macro-botanical and phytolith analyses of the Guijiabao site show that millet was the major crop, similar to the findings at the Yingpanshan site in Sichuan [@Zhao2011]. However, plant remains from many sites in southwest China reveal a predominance of rice, such as at the Gaoshan and Baodun sites in Sichuan [@JiangGaoshan2017] and Baiyangcun sites in Yunnan [@Jiang2016; @Henglanshan2014]. This indicates that there were variations in agricultural practices in southwest China. To better understand the extent of the adoption of new crops, organic residue analysis of ceramics is a useful approach since we can directly study food residues related to culinary practices, including food preparation, processing, and cooking. This provides another way to examine the expansion of agriculture in the Yanyuan area.

@@

<!--discussion about possible ruminant fat origin-->
The ruminant fats identified in the Neolithic potsherds (ca. 4500 BP) from the Guijiabao and Gaoshan sites could have been derived from deer, sheep, or horses that were fed on a C3 diet, based on their isotopic values that are consistent with global references [@Evershed2002; @Mileto2017]. This finding, especially the consumption of deer, also corresponds with unearthed faunal remains from the sites and the local natural environments. Stone artifacts found at the site, including stone balls and arrowheads, further indicate hunting activities [@Hao2022; @Guijiabao2020a]. However, whether horse was part of local food choices in southwest China around 4500 years ago is doubtful. First, the Pleistocene and Neolithic sites having wild horse bones are primarily concentrated in northern China, with only sporadic sites in the south [@Yuan2006]. Second, domesticated horses first found in southwest China date to the Dian culture (4th century BC to 1st century AD) in eastern Yunnan [@Lei2009; @Chiou2004], which is 3000 years later than the Neolithic period that we are focusing on. Thus, supported by faunal evidence, deer was most likely the predominant local food choices. It is also possible that local wild bovids, such as yaks or water buffalo, were food sources during this period, as these animals have been found in sites in southwest China [@Zhang2016; @Chen2020].

@@

In general, based on our results, ruminant fats from local wild animals, such as deer, were the predominant food source during the Neolithic period for peoples at Guijiabao. A subsistence that relied on hunting to a certain degree could be plausible, given the abundance of wild animals in the natural settings of the Hengduan mountain ranges, as indicated by the faunal remains from the Yingpanshan site [@He2020]. Crop cultivation could have been adopted as a supplement to local food sources when it was introduced in the Neolithic period, but crops might have only accounted for a small proportion of the food sources. A previous study using isotopic analysis of human bones suggests that the diet at the Guijiabao site was a C4-based diet [@Lin2022], but we only identified a signal from C3 plants from one pottery sample without any signal from C4 plants. This inconsistency could have resulted from different sampling strategies or from the sample size issue, where our results reflect only a small portion of food content. Nevertheless, our results suggest that farming practices during the Neolithic period might not have been adopted as much as we previously thought. An approach analyzing organic residues may only provide a narrow perspective on local diets, but it could improve our current understanding of food practices through presence of food residues that were cooked in ceramic vessels. A large-scale organic residue analysis of ceramics across different pottery forms in a wider region would offer insight into temporal and regional variations in the local diet during the Neolithic period in southwest China.

# Conclusion
Our case study on organic residues in pottery fragments from the sites at Guijiabao and Gaoshan in southwest China reveals the presence of lipids that can be analyzed to determine food sources. The foods cooked in the ceramics included ruminant fats and terrestrial plants during the Neolithic period, when millet and rice spread into southwest China. The source of terrestrial plants could have been C3 plants, implying the cooking of rice or other wild weeds in the pots. However, the evidence of direct cooking of either C3 or C4 plants shows that the amount of C3/C4 plants cooked is not in proportion to the large quantity of plant remains found at the Guijiabao site, especially C4 plants such as millet. Instead, ruminant fats from local animals, such as wild deer, could have been a common food source during the Neolithic period. This suggests that hunting might have been an important part of local subsistence strategy, supplemented by rice/millet farming when they were introduced. Our results from the Guijiabao site could represent a subsistence strategy for adapting to the natural environment of the Hengduan Mountains. We recognize the limitations of our interpretation of the local diet due to the small sample size. Nevertheless, our study could offer an approach using organic residue analysis to determine food choices and explore the extent of the adoption of farming when rice/millet agriculture first spread into the Yanyuan area during the Neolithic period. A more systematic sampling of potsherds across different phases and pottery types may give a clear picture of diet and subsistence in southwest China during the Neolithic period of transition.

>>>>

---
title: "Investigating Neolithic pottery use from Guijiabao in the Yanyuan Basin in southwestern China using organic residue analysis"
author:
  - Li-Ying Wang:
      email: liyingw@asihp.net
      institute: [AcS]
      correspondence: true
  - Kuei-chen Lin:
      email: kueichenlin@asihp.net
      institute: [AcS]
      correspondence: false
  - Zhiqing Zhou:
      email: zhouzhiQ2000@163.com
      institute: [InsC]
      correspondence: false
institute:
  - AcS: Institute of History and Philology, Academia Sinica 
  - InsC: Chengdu Institute of Cultural Relics and Archaeology
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
    The spread of new crops or terrestrial animals across areas in southwestern China during the Neolithic period is one of the important archaeological issues to understand the process of adoption of new culinary practices in this region. It is widely accepted that rice was introduced into southwest China in 4700 BP at the latest and millets came even earlier. However, to what extent new crops were used by local peoples at that time remains unclear. The excavation of Guijiabao in the Yanyuan basin in southwestern China uncovered abundant materials with well-preserved stratigraphic records and has provided a unique opportunity to understand the local adoption of new crops. To further understand pottery use and food practices at Guijiabao, we investigated organic residues preserved in potsherds using lipid biomarker identification and compounds-specific stable isotope analysis. The results show signals of terrestrial animals in most potsherds and a C3 plant source in one potshard, during the Neolithic period. Our case study indicates that hunting might have been the primary means for obtaining subsistence, supplemented by rice/millet farming. This study also demonstrates an approach using organic residue analysis to explore the extent of the adoption of farming during the Neolithic period in southwestern China.

keywords: |
  Southwestern China; Neolithic period; Lipid residue analysis; Stable isotope analysis; Ceramics; Culinary practices
highlights: |
  These are the highlights. 
---

@@

# Introduction
<!-- intro to the topic and research aim-->
Movements of people between geographically distinct areas are usually accompanied by the exchange of goods, technology, or ideas that could have cultural or economic impacts on local people groups involved in the interactions [@Burmeister2000]. The Hengduan Mountain ranges and valleys in southwestern China are geographical areas that connect the Qinghai-Tibetan Plateau and the Yunnan–Guizhou Plateau and by which movement to these plateaus must occur. They play an important role in the cultural exchanges and spread of crops into southwestern China and even further to mainland Southeast Asia [@Ma2021; @Lin2022; @Hao2022; @Tong1986]. Despite a long-standing debate on the routes of rice/millet dispersal from northwest China and the middle Yangtze valley to southwestern China [@Guedes2013; @DAlpoim2018; @Liu2022], it is widely accepted that mixed farming was practiced in southwestern Sichuan around 4700 BP and in western Yunnan before 4600 BP [@Jiang2016; @YunnanMuseum1981; @Hao2022; @Zhang2010; @Dal2022; @Chen2015]. The Hengduan Mountain regions in the Neolithic period are therefore essential for understanding the processes of agricultural expansion. Recent archaeological evidence from the Yanyuan Basin in the middle Hengduan Mountains further reveals that millet and rice were present at roughly the same period as the grains found in Sichuan [@Huan2022; @Hao2022]. However, to what extent the two crops were incorporated into the local diet is still not fully understood.

@@

```{r SW-china-map, fig.cap = "The archaeological sites Guijiabao and Gaoshan and other locations in southwestern China that are mentioned in this study. Map data is from naturalearthdata.com"}
library(here)
knitr::include_graphics(here("analysis", "figures", "Sichuan-sites-map.png"))
```

@@

Although the recent findings at the Guijiabao and Gaoshan sites support mixed agriculture as the primary subsistence strategy, it is still unclear how much millet or rice was included in the local diets during the Neolithic period. The macro-botanical and phytolith analyses of the Guijiabao site show that millet was the major crop, similar to the findings at the Yingpanshan site in Sichuan [@Zhao2011]. However, plant remains from many sites in southwestern China reveal a predominance of rice, such as at the Gaoshan and Baodun sites in Sichuan [@JiangGaoshan2017] and Baiyangcun sites in Yunnan [@Jiang2016; @Henglanshan2014]. This indicates that there were variations in agricultural practices in southwestern China. To better understand the extent of the adoption of new crops, organic residue analysis of ceramics is a useful approach since we can directly study food residues related to culinary practices, including food preparation, processing, and cooking. This provides another way to examine the expansion of agriculture in the Yanyuan area.

@@

<!--discussion about possible ruminant fat origin-->
The ruminant fats identified in the Neolithic potsherds (ca. 4500 BP) from the Guijiabao and Gaoshan sites could have been derived from deer, sheep, or horses that were fed on a C3 diet, based on their isotopic values that are consistent with global references [@Evershed2002; @Mileto2017]. This finding, especially the consumption of deer, also corresponds with unearthed faunal remains from the sites and the local natural environments. Stone artifacts found at the site, including stone balls and arrowheads, further indicate hunting activities [@Hao2022; @Guijiabao2020a]. However, whether horse was part of local food choices in southwestern China around 4500 years ago is doubtful. First, the Pleistocene and Neolithic sites having wild horse bones are primarily concentrated in northern China, with only sporadic sites in the south [@Yuan2006]. Second, domesticated horses first found in southwestern China date to the Dian culture (4th century BC to 1st century AD) in eastern Yunnan [@Lei2009; @Chiou2004], which is 3000 years later than the Neolithic period that we are focusing on. Thus, supported by faunal evidence, deer was most likely the predominant local food choices. It is also possible that local wild bovids, such as yaks or water buffalo, were food sources during this period, as these animals have been found in sites in southwestern China [@Zhang2016; @Chen2020].

@@

In general, based on our results, ruminant fats from local wild animals, such as deer, were the predominant food source during the Neolithic period for peoples at Guijiabao. A subsistence that relied on hunting to a certain degree could be plausible, given the abundance of wild animals in the natural settings of the Hengduan mountain ranges, as indicated by the faunal remains from the Yingpanshan site [@He2020]. Crop cultivation could have been adopted as a supplement to local food sources when it was introduced in the Neolithic period, but crops might have only accounted for a small proportion of the food sources. A previous study using isotopic analysis of human bones suggests that the diet at the Guijiabao site was a C4-based diet [@Lin2022], but we only identified a signal from C3 plants from one pottery sample without any signal from C4 plants. This inconsistency could have resulted from different sampling strategies or from the sample size issue, where our results reflect only a small portion of food content. Nevertheless, our results suggest that farming practices during the Neolithic period might not have been adopted as much as we previously thought. An approach analyzing organic residues may only provide a narrow perspective on local diets, but it could improve our current understanding of food practices through presence of food residues that were cooked in ceramic vessels. A large-scale organic residue analysis of ceramics across different pottery forms in a wider region would offer insight into temporal and regional variations in the local diet during the Neolithic period in southwestern China.

# Conclusion
Our case study on organic residues in pottery fragments from the sites at Guijiabao and Gaoshan in southwestern China reveals the presence of lipids that can be analyzed to determine food sources. The foods cooked in the ceramics included ruminant fats and terrestrial plants during the Neolithic period, when millet and rice spread into southwestern China. The source of terrestrial plants could have been C3 plants, implying the cooking of rice or other wild weeds in the pots. However, the evidence of direct cooking of either C3 or C4 plants shows that the amount of C3/C4 plants cooked is not in proportion to the large quantity of plant remains found at the Guijiabao site, especially C4 plants such as millet. Instead, ruminant fats from local animals, such as wild deer, could have been a common food source during the Neolithic period. This suggests that hunting might have been an important part of local subsistence strategy, supplemented by rice/millet farming when they were introduced. Our results from the Guijiabao site could represent a subsistence strategy for adapting to the natural environment of the Hengduan Mountains. We recognize the limitations of our interpretation of the local diet due to the small sample size. Nevertheless, our study could offer an approach using organic residue analysis to determine food choices and explore the extent of the adoption of farming when rice/millet agriculture first spread into the Yanyuan area during the Neolithic period. A more systematic sampling of potsherds across different phases and pottery types may give a clear picture of diet and subsistence in southwestern China during the Neolithic period of transition.
